\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-14 k-planar.tex
added new  map graph lemma

\documentclass{patmorin}
\listfiles
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
%\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}
\usepackage{thm-restate}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

@@

%\usepackage{doi} % To make plainnat handle doi's properly

% \usepackage[mathlines]{lineno}
% \setlength{\linenumbersep}{2em}
% \linenumbers
% \rightlinenumbers
% \linenumbers
% \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
%  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
%  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
%  \renewenvironment{#1}%
%     {\linenomath\csname old#1\endcsname}%
%     {\csname oldend#1\endcsname\endlinenomath}}%
% \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
%  \patchAmsMathEnvironmentForLineno{#1}%
%  \patchAmsMathEnvironmentForLineno{#1*}}%
% \AtBeginDocument{%
% \patchBothAmsMathEnvironmentsForLineno{equation}%
% \patchBothAmsMathEnvironmentsForLineno{align}%
% \patchBothAmsMathEnvironmentsForLineno{flalign}%
% \patchBothAmsMathEnvironmentsForLineno{alignat}%
% \patchBothAmsMathEnvironmentsForLineno{gather}%
% \patchBothAmsMathEnvironmentsForLineno{multline}%
% }
%
%
% \allowdisplaybreaks
%\sloppy
%%%%%%%%%%
 \makeatletter
 \def\NAT@spacechar{~}
 \makeatother

@@

\note{PM}{The trick I used (using the thm counter everywhere) to get consecutive Theorem, Lemma, etc numbering has broken cleverref.  Everything is now references as a Theorem.  Find a solution that works.}
% \note{DW}{I suggest we use consecutive numbering of theorems, lemmas, etc. Theorem 1, Lemma 2, Observation 3, etc.}

% \note{PM}{Agreed. Done.}

@@

\begin{thm}
\label{kPlanarProduct}
Every $k$-planar graph is contained in $H\boxtimes P\boxtimes K_{18k^2+48k+30}$, for some graph $H$ of treewidth $\binom{k+4}{3}-1$ and for some path $P$.
\end{thm}

@@

For any integer $d\ge 3$, a (multi)graph $G$ is \defin{$d$-framed} if it admits a drawing in the plane having a biconnected plane skeleton (multi)graph $G_0$, whose faces have size at most $d$, and whose crossing edges are in the interior of these faces.  This definition (for simple graphs) was introduced by \citet{BDGGMR}. Every $d$-framed graph can be described by a $(2,\binom{d}{2})$-shortcut system applied to a planar graph. Thus \cref{ShortcutProduct} is applicable. However, for framed graphs we prove the following stronger result, where the treewidth bound on $H$ is best possible.

\begin{thm}\label{d_framed_product_stucture}
Every $d$-framed graph is contained in $H\boxtimes P\boxtimes K_{d+3\floor{d/2}-3}$ for some planar graph $H$ with treewidth at most 3 and for some path $P$.
\end{thm}

This result is used below to obtain product structure theorems for 1-planar graphs and for map graphs. For 1-planar graphs it is essential that we allow skeletons with parallel edges.

@@

\begin{thm}
\label{1-planar}
Every 1-planar graph is contained in $H\boxtimes P\boxtimes K_{7}$ for some planar graph $H$ with treewidth at most 3 and for some path $P$.
\end{thm}

@@

\cref{1-planar} is an immediate consequence of \cref{d_framed_product_stucture} and the next lemma, which is of independent interest. We assume that no two edges incident to a common vertex of the $1$-plane graph $G$ cross each other since, in a 1-plane graph, such a crossing can always be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross.\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}  % We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.

@@

\begin{proof}
Let $G$ be a $1$-plane graph.  We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If two vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no two such edges will appear on the boundary of a common face.

@@

\note{DW}{I am pretty sure one can get treewidth 3 for $(g,1)$-planar graphs, combining \cref{framed} with the treewidth 3 result in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). Okay?}

@@

Map graphs are defined as follows. Start with a graph $G_0$ embedded in a surface of Euler genus $g$, with each face labelled a `nation' or a `lake', where each vertex of $G_0$ is incident with at most $d$ nations. Let $G$ be the graph whose vertices are the nations of $G_0$, where two vertices are adjacent in $G$ if the corresponding faces in $G_0$ share a vertex. Then $G$ is called a \defin{$(g,d)$-map graph}.  A $(0,d)$-map graph is called a (plane) \defin{$d$-map graph}; see \citep{FLS-SODA12,CGP02} for example.
The $(g,3)$-map graphs are precisely the graphs of Euler genus at most $g$; see \citep{dujmovic.eppstein.ea:structure}. So $(g,d)$-map graphs generalise graphs embedded in a surface, and we now assume that $d\geq 4$ for the remainder of this section.

@@

\note{DW}{I am pretty sure one can get treewidth 3 here using \cref{framed} and the treewidth 3 result in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). Okay?}

@@

\begin{lem}
Every $d$-map graph is a subgraph of $\widehat{G}$ for some 2-connected plane graph $G$ in which every facial cycle has length in $\{3,\dots,d\}$.
\end{lem}

\begin{cor}
\label{NewMapGraph}
Every $d$-map graph is contained in $H \boxtimes P \boxtimes K_{ d + 3\floor{d/2} -3 }$ for some graph $H$ with $\tw(H) \leq 3$ and for some path $P$.
\end{cor}

>>>>

\documentclass{patmorin}
\listfiles
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{paralist}
\usepackage{stmaryrd}
\usepackage{thm-restate}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

@@

% \allowdisplaybreaks
%\sloppy
%%%%%%%%%%
 \makeatletter
 \def\NAT@spacechar{~}
 \makeatother

@@

@@

\begin{thm}
\label{kPlanarProduct}
Every $k$-planar graph is contained in $H\boxtimes P\boxtimes K_{18k^2+48k+30}$, for some graph $H$ of treewidth $\binom{k+4}{3}-1$ and for some path $P$.
\note{DW}{Do we want to mention ``every $1$-planar graph is contained in $H\boxtimes P\boxtimes K_7$ for some planar graph $H$ of treewidth $3$ and for some path $P$'' here?}
\end{thm}

@@

For any integer $d\ge 3$, a (multi)graph $G$ is \defin{$d$-framed} if it admits a drawing in the plane having a biconnected plane (multi) subgraph $G_0$, whose facial cycles have length at most $d$,  and whose crossing edges are in the interior of these faces \note{DW}{I suggest we replace ``and whose crossing edges are in the interior of these faces'' by ``such that each edge in $E(G)\setminus E(G_0)$ is in the interior of some face of $G_0$. } The graph $G_0$ is called the \defin{frame} of $G_0$. This definition (for simple graphs) was introduced by \citet{BDGGMR}. Every $d$-framed multigraph can be described by a $(2,\binom{d}{2})$-shortcut system applied to a planar graph. Thus \cref{ShortcutProduct} is applicable. However, for framed graphs we prove the following stronger result, where the treewidth bound on $H$ is best possible\footnote{A multigraph $G$ is \defin{contained} in a graph $G'$ if the simple graph underlying $G$ is isomorphic to a subgraph of $G'$.}.

\begin{thm}\label{d_framed_product_stucture}
Every $d$-framed multigraph is contained in $H\boxtimes P\boxtimes K_{d+3\floor{d/2}-3}$ for some planar graph $H$ with treewidth at most 3 and for some path $P$.
\end{thm}

This result is used below to obtain product structure theorems for 1-planar multigraphs and for map graphs. For simple 1-planar graphs it is essential that we allow frames with parallel edges.

@@

\begin{thm}
\label{1-planar}
Every 1-planar graph is contained in $H\boxtimes P\boxtimes K_7$ for some planar graph $H$ with treewidth at most 3 and for some path $P$.
\end{thm}

@@

\cref{1-planar} is an immediate consequence of \cref{d_framed_product_stucture} and the next lemma, which is of independent interest.

@@

\begin{proof}
Let $G$ be a $1$-plane graph. We may assume that no two edges incident to a common vertex of $G$ cross, since such a crossing can be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}.
  % We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.
We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If two vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no two such edges will appear on the boundary of a common face.

@@

\note{DW}{I am pretty sure one can get treewidth 3 for $(g,1)$-planar graphs, combining \cref{framed} with the method in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). Okay?}

@@

\note{DW}{This subsection needs a clean-up; define $(g,d)$-framed graph. Is the following lemma known (at least in the $g=0$ case)?}

\begin{lem}
\label{MewMapGraphLemma}
for all integers $g\geq 0$ and $d\geq 3$, every $(g,d)$-map graph is a spanning subgraph of  $(g,d)$-framed graph.
\end{lem}

\begin{proof}
Let $G_0$ be a graph embedded in a surface of Euler genus $g$, with each face labelled a nation or a face, and where each vertex of $G_0$ is incident with at most $d$ nations. If two lakes have a common edge in their boundaries, then we may delete the edges and merge the lakes. Now assume that no two lakes have a common edge in their boundaries.
Let $G$ be the corresponding map graph. 
Let $G_1$ be the graph with $V(G_1):=V(G)$, which is the set of nations. 
Consider vertices $v$ and $w$ of $G_1$. 
Let $A$ and $B$ be the nation faces of $G_0$ corresponding to $v$ and $w$.
Then $vw\in E(G_1)$ if $A$ and $B$ have a common edge in their boundaries, 
or there is a vertex $x$ of $G_0$ in the boundary of $A$ and $B$, and in the clockwise or anticlockwise order of faces around $x$, all the faces between $A$ and $B$ are lakes. 
Then $G_1$ has Euler genus at most $g$. 
The faces of $G_1$ correspond to vertices of $G_0$ and lakes of $G_0$.
Triangulate each face of $G_1$ corresponding to a lake of $G_0$. For each vertex 
$x$ of $G_0$ the size of the face of $G_1$ corresponding to $x$ equals the number of nations incident to $x$, which is at most $d$. 
So every face of $G_1$ has size at most $d$. 
If $vw\in E(G)$ then the nation faces of $G_0$ corresponding to $v$ and $w$ have a vertex $x$ of $G_0$ in common, implying that  $v$ and $w$ are both on the face of $G_1$ corresponding to $x$, and $vw\in E(\widehat{G_1})$. Hence $G\subseteq \widehat{G_1}$.	
\end{proof}

\begin{thm}
\label{NewMapGraph}
Every $d$-map graph is contained in $H \boxtimes P \boxtimes K_{ d + 3\floor{d/2} -3 }$ for some graph $H$ with $\tw(H) \leq 3$ and for some path $P$.
\end{thm}

Map graphs are defined as follows. Start with a graph $G_0$ embedded in a surface of Euler genus $g$, with each face labelled a `nation' or a `lake', where each vertex of $G_0$ is incident with at most $d$ nations. Let $G$ be the graph whose vertices are the nations of $G_0$, where two vertices are adjacent in $G$ if the corresponding faces in $G_0$ share a vertex. Then $G$ is called a \defin{$(g,d)$-map graph}.  A $(0,d)$-map graph is called a (plane) \defin{$d$-map graph}; see \citep{FLS-SODA12,CGP02} for example.
The $(g,3)$-map graphs are precisely the graphs of Euler genus at most $g$; see \citep{dujmovic.eppstein.ea:structure}. 
%So $(g,d)$-map graphs generalise graphs embedded in a surface, and we now assume that $d\geq 4$ for the remainder of this section.

@@

\note{DW}{I am pretty sure one can get treewidth 3 here using \cref{framed} and the method in \citep{DHHW}. I think the best place for this result is reference \citep{DHHW} since the present paper pre-dates \citep{DHHW} (and this addition will increase the significance of \citep{DHHW}). Okay?}

@@
